---
title: "Lab 6 - Hierarchical Models"
author: "Arun Oakley-Cogan"
date: "2024-11-02"
output: html_document
---

```{r}
library(rethinking)
rm(list=ls())
```

Today we are going to look at Hierarchical\\MultiLevel\\Varying Intercepts\\Partial Pooling\\Random Effects models, and we are going to compare them with their traditional fixed-effects model counterparts we have been focusing on throughout this course.

**Tadpole example**

```{r}
data("reedfrogs")
tdata <- reedfrogs
head(tdata)
```

Fixed Effects Model

$$
\large{
\begin{align*}
Survival_i \sim Binomial(Density_i, p_i)\\
logit(p_i) = alpha_{tank[i]}\\
alpha_{tank[i]} \sim Normal(?, ?)\\
\end{align*}
}
$$

A fixed-effects model, each tank is modelled independently. The model will iterate through tank, extracting information about the survival in that tank, and using that to update the probability of survival. Once the model has learned all it can from a tank it moves onto the next one, forgetting everything it has learned about previous tanks, and starts again from scratch.

This is SUPER inefficient, because yes, while each tank is actually different, they are also alike, so when it learns about the regular features of one tank, it can use that information to help it learn about a new tank quicker.

```{r}
# create our categorical index variables
tdata$tank <-  1:nrow(tdata)
head(tdata)
```

**Prior predictive simulation**

```{r}
# What do we think the average tank survival rate should be?
N <- 1000 # number of samples
tank_prior_1 <- rnorm(N, 0.5, 0.1 )

plot(density(tank_prior_1), xlim=c(-1,1), xlab="Log-Odds Scale", main="Prior predictive plot")
plot(density(inv_logit(tank_prior_1)), xlim=c(0,1), xlab="Probability Scale", main="Prior predictive plot")

```

```{r}
# let's try and get that prior to be 50% on the OUTCOME scale.
# starting off with a N(0,1)
tank_prior_2 <- rnorm(N, 0, 1)
plot(density(tank_prior_2), xlab="Log-Odds Scale", main="Prior predictive plot")
plot(density(inv_logit(tank_prior_2)), xlim=c(0,1), xlab="Probability Scale", main="Prior predictive plot")
```

```{r}
# that's actually not too bad, we could make this a little more regularizing (skeptical) of extreme values
# I'll tighten this up by decreasing the variance.
tank_prior_3 <- rnorm(N, 0, 0.6)
plot(density(tank_prior_3), xlab="Log-Odds Scale", main="Prior predictive plot", col="deepskyblue4")
lines(density(tank_prior_2), col="orange")
legend("topright", legend=c("Tank N(0,0.75) Prior", "Tank N(0,0.1) Prior"), col=c("deepskyblue4", "orange"), lty=c(1,1))
plot(density(inv_logit(tank_prior_3)), xlim=c(0,1), xlab="Probability Scale", main="Prior predictive plot",  col="deepskyblue4")
lines(density(inv_logit(tank_prior_2)), , col="orange")
legend("topright", legend=c("Tank N(0,0.75) Prior", "Tank N(0,1) Prior"), col=c("deepskyblue4", "orange"), lty=c(1,1))
```

**Data setup**

```{r}
list_tad <- list(
  survival = tdata$surv,
  density = tdata$density,
  tank = tdata$tank
)
```

**Model setup**

```{r}
model_tad <- ulam(
  alist(
    survival ~ dbinom(density, p), # data likelihood
    logit(p) <- alpha[tank],
    alpha[tank] ~ dnorm(0,0.6)
  ), data=list_tad, chains=4, log_lik = TRUE
)
```

Let's take our first look at how the model is doing ..

**Posterior Estimates & Diagnostics**

```{r}
precis(model_tad, depth=2)
traceplot(model_tad)
```

**Posterior Predictions**

```{r}
plot(precis(model_tad, depth=2))
posterior <- extract.samples(model_tad)
mean_surv <- inv_logit(apply(posterior$alpha, 2, mean))

plot( tdata$propsurv , ylim=c(0,1) , pch=16 , xaxt="n" , xlab="tank" , ylab="proportion survival" , col=1 ) 
axis( 1 , at=c(1,16,32,48) , labels=c(1,16,32,48) )
points(mean_surv, col=4)
abline( h=mean(tdata$propsurv) , lty=2, col=1 )
abline( h=mean(mean_surv) , lty=2, col=4 )
#abline( h=mean(inv_logit(tank_prior)) , lty=1, col=4 )
legend("bottomleft", legend = c("Observed Proportions", "Fixed-Effect Proportions"), col = c(1,4), pch = c(16, 1))
```

Hierarchical Model, Bayes Notation

These models do not to forget about each tank, as it both models the observations of each tank, but also models the populations of tanks at the same time. Doing so allows the model to have some memory of each tank.

$$
\large{
\begin{align*}
Survival_i \sim Binomial(Density_i, p_i)\\
logit(p_i) = alpha_{tank[i]}\\
\alpha_{tank[i]} \sim Normal(a\_bar, \sigma)\\
a\_bar \sim Normal(0, 1.5)\\
\sigma \sim Exp(1)\\
\end{align*}
}
$$

**Prior predictive simulation**

```{r}
N <- 1000 # number of samples
a_bar_prior <- rnorm(N, 0, 1.5)
sigma_prior <- rexp(N, 1)
tank_prior <- rnorm(N, a_bar_prior, sigma_prior)

plot(density(tank_prior), xlim=c(-7,7), ylim=c(0,1), xlab="Log-Odds Scale", main="Prior predictive plot")
legend("topright", legend=c("Tank Survival Prior"), col=c("black"), lty=c(1,1))
plot(density(inv_logit(tank_prior)), xlim=c(0,1), ylim=c(0,4), xlab="Probability Scale", main="Prior predictive plot")
legend("topright", legend=c("Tank Prior"), col=c("black"), lty=c(1,1))
```

**Model setup**

```{r}
model_tad_h <- ulam(
  alist(
    survival ~ dbinom(density, p), # data likelihood
    logit(p) <- alpha[tank],
    alpha[tank] ~ dnorm(a_bar,sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ), data=list_tad, chains=4, log_lik = TRUE
)
```

Model Comparison Visual

```{r}
posterior_h <- extract.samples(model_tad_h)
mean_surv <- inv_logit(apply(posterior$alpha, 2, mean))
mean_surv_h <- inv_logit(apply(posterior_h$alpha, 2, mean))

plot( tdata$propsurv , ylim=c(0,1) , pch=16 , xaxt="n" , xlab="tank" , ylab="proportion survival" , col=1 ) 
axis( 1 , at=c(1,16,32,48) , labels=c(1,16,32,48) )
points(mean_surv, col=4)
points(mean_surv_h, col=2)
abline( h=mean(inv_logit(posterior_h$a_bar)) , lty=2, col=2 )
abline( h=mean(mean_surv) , lty=2, col=4 )
legend("bottomleft", legend = c("Observed Proportions", "Fixed-Effect Proportions","Hierarchy Proportions"), col = c(1,4,2), pch = c(16, 1, 1))
```

Where shrinkage is stronger, the further a tanks raw proportion is from the global average

```{r}
compare( model_tad , model_tad_h )
```

Let's add in another covariate and see if we can better model the survival. 

Fixed Effects Model

$$
\large{
\begin{align*}
Survival_i \sim Binomial(Density_i, p_i)\\
logit(p_i) = alpha_{tank[i]} + betaP*Pred\\
alpha_{tank[i]} \sim Normal(0, 0.6)\\
betaP_{tank[i]} \sim Normal(?, ?)\\
\end{align*}
}
$$

```{r}
# create a dummy/binary variable for predators in a tank
tdata$pred_id <-  ifelse(tdata$pred == "no", 0L, 1L)
```

**Prior predictive simulation**

```{r}
# So here betaP is the DIFFERENCE in average survival for tanks that have a predator. 
# I assume this will be a negative difference
N <- 1000 # number of samples
# prior plots
tank_prior <- rnorm(N, 0, 0.6 )
pred_prior <- rnorm(N,-0.5, 1)
plot(density(tank_prior), xlab="Log-Odds Scale", main="Prior plot", col="deepskyblue4")
lines(density(pred_prior), col="orange")
plot(density(inv_logit(tank_prior)), xlim=c(0,1), xlab="Probability Scale", main="Prior plot", col="deepskyblue4")
lines(density(inv_logit(pred_prior)), col="orange")
```

```{r}
# prior predictive plots
plot(density(tank_prior + pred_prior), xlab="Log-Odds Scale", main="Prior Predictive plot")
plot(density(inv_logit(tank_prior + pred_prior)), xlim=c(0,1), xlab="Probability Scale", main="Prior Predictive plot")
```

**Data setup**

```{r}
list_tad_p <- list(
  survival = tdata$surv,
  density = tdata$density,
  pred_id = as.integer(tdata$pred_id),
  tank = tdata$tank,
  pred = as.integer(tdata$pred_id) + 1
)
```

**Model setup**

```{r}
model_tad_p <- ulam(
  alist(
    survival ~ dbinom(density, p), # data likelihood
    logit(p) <- alpha[tank] + betaP*pred_id,
    alpha[tank] ~ dnorm(0,0.6),
    betaP ~ dnorm(-0.5, 1)
  ), data=list_tad_p, chains=4, log_lik = TRUE
)
```

```{r}
precis(model_tad_p, depth=2)
```

**Posterior Predictions**

```{r}
posterior_p <- extract.samples(model_tad_p)
mean_surv_p <- link(model_tad_p)
mu_mean_surv_p <- apply(mean_surv_p, 2, mean)
plot( tdata$propsurv , ylim=c(0,1) , pch=16 , xaxt="n" , xlab="tank" , ylab="proportion survival" , col=1 ) 
axis( 1 , at=c(1,16,32,48) , labels=c(1,16,32,48) )
points(mu_mean_surv_p, col="orange")
points(mean_surv, col="deepskyblue4")
abline( h=mean(tdata$propsurv) , lty=2, col=1 )
abline( h=mean(mean_surv_p) , lty=2, col="orange" )
legend("bottomleft", legend = c("Observed Proportions", "Fixed-Effect Predators", "Fixed-Effect No Pred"), col = c(1,"orange", "deepskyblue4"), pch = c(16, 1, 1))
```

$$
\large{
\begin{align*}
Survival_i \sim Binomial(Density_i, p_i)\\
logit(p_i) = \alpha_{tank[i]} + betaP*Pred\\
\alpha_{tank[i]} \sim Normal(a\_bar, \sigma)\\
a\_bar \sim Normal(0, 1.5)\\
\sigma \sim Exp(1)\\
betaP \sim Normal(-0.5,1)
\end{align*}
}
$$

```{r}
model_tad_ph <- ulam(
  alist(
    survival ~ dbinom(density, p), # data likelihood
    logit(p) <- alpha[tank] + betaP*pred_id,
    alpha[tank] ~ dnorm(a_bar,sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1),
    betaP ~ dnorm(-0.5,1)
  ), data=list_tad_p, chains=4, log_lik = TRUE
)
```

```{r}
posterior_hp <- extract.samples(model_tad_ph)
mean_surv_hp <- link(model_tad_ph)
mu_mean_surv_hp <- apply(mean_surv_hp, 2, mean)
plot( tdata$propsurv , ylim=c(0,1) , pch=16 , xaxt="n" , xlab="tank" , ylab="proportion survival" , col=1 ) 
axis( 1 , at=c(1,16,32,48) , labels=c(1,16,32,48) )
points(mu_mean_surv_hp, col="red")
abline( h=mean(inv_logit(posterior_hp$a_bar)) , lty=2, col="red" )
abline( h=mean(inv_logit(posterior_hp$a_bar + posterior_hp$betaP)) , lty=2, col="green" )
legend("bottomleft", legend = c("Observed Survival","Hierarchy Survival", 'Heirarchy w/ Pred Survival'), col = c(1,"orange", "red"), pch = c(16, 1, 1))
```

```{r}
plot(coeftab(model_tad_h, model_tad_ph), pars=c("sigma", "betaP", "a_bar"))
```
```{r}
compare(model_tad, model_tad_h, model_tad_p, model_tad_ph)
```
```{r}

```
