---
title: "Lab 1, Our first model."
author: "Arun Oakley-Cogan"
date: "2024-09-04"
output: html_document
---

```{r message = F}
library(tidyverse)
library(rethinking)
```

### Our first model. Estimating the presence/absence of frogs at a pond.

**Scenario**: We go to a pond 30 times and record the presence (1) or absence (0) of frogs, and estimate the probability of seeing a frog when visiting the pond.

### Generative model and simulated data.

```{r}
# number of trips to pond
n_trips <- 30

# the true probability of seeing a frog in this pond
true_prob <- 0.35

# simulate data from generative model
sim_data <- rbinom(n_trips, size = 1, prob = true_prob)

# data for model
data_list <- list(
  trips = length(sim_data), # total number of trips
  frogs = sum(sim_data) # number of times seen frogs
)
sim_data;
data_list
```

### Build statistical model: grid approximation

**Model Components**

-   **Grid**, a finite set of parameters values to be used
-   **Prior Distribution**, prior probability for each value of the parameter
-   **Data Likelihood**, the distribution function assigned to our observed variables. This function gives us the likelihood of the data for every plausible value of our parameter.
-   **Posterior Distribution**, contains the probability of different parameter values, conditional on the data and model.

#### Bayesian Notation

$$F \sim Binomial(n, p)$$ $$p \sim Uniform(0,1)$$

```{r}
# grid approximation method
# define our parameter grid
par_grid <- seq(from=0, to=1, length.out=20)

# define a uniform prior
prior <- rep(1, length(par_grid))
prior <- prior / sum(prior)

# estimate the likelihood of seeing frogs in this many trips
likelihood <- dbinom(data_list$frogs, data_list$trips, prob = par_grid)

# estimate posterior
posterior <- (likelihood * prior) / sum(likelihood * prior)

# plot posterior distribution
plot(par_grid, posterior, col='blue', type='b' , xlab='Probability of seeing frogs', ylab='Density')
# plot prior distribution
points(x = par_grid, y = prior, type='b', col='orange')
abline(v = 0.3, col="red")
legend("topright", legend=c("Posterior", "Prior", "True Probability"),
       col=c("blue", "orange", "red"), lty=1)

```

```{r}
# summarise our posterior - grid approximation
samples <- sample( par_grid , prob=posterior , size=1000 , replace=TRUE )

# mean
mean(samples)

# how much of the posterior is above/below a value?
sum( posterior[ par_grid < 0.4 ] )
sum( posterior[ par_grid > 0.4 ] )

# compatibility intervals
# percentile intervals (PI) - even distribution above/below interval, indicate shape of distribution
PI(samples, prob=0.5)
# highest posterior density interval (HPDI) - best representatives of parameter values
HPDI(samples, prob=0.5)

```

### Build statistical model: MCMC (ulam)

```
a.  **Model data,** a list containing our variables and data
b.  **Prior Distribution**, prior probability of the parameter at each probability
c.  **Data Likelihood**, the distribution function assigned to our observed variables. This function gives us the likelihood of the data for every plausible value of our parameter.
```

```{r}
# model data
data_list
```

```{r}
# MCMC method using ulam with uninformed prior
frog_model <- ulam(
  alist(
    frogs ~ dbinom(trips, p), # data likelihood
    p ~ dunif(0, 1) # p prior
  ), data = data_list
)

```

```{r}
# model summary
precis(frog_model)
```

```{r}
# posterior samples
samples <- extract.samples(frog_model, n=10000)

# plots
curve(dnorm(x, mean(samples$p), sd(samples$p)), xlab="Probability of seeing frogs", ylab="Density", col="blue")
curve((dunif(x, 0, 1)), add = TRUE, col="orange")
abline(v = 0.35, col="red")
legend("topright", legend=c("Posterior", "Prior", "True Probability"),
       col=c("blue", "orange", "red"), lty=1)
```

```{r}
# posterior summaries
# above 30%
sum(samples$p > 0.35) / length(samples$p)

# compatibility intervals
# percentage interval
PI(samples$p, prob=0.5)
# highest posterior density interval
HPDI(samples$p, prob=0.96)
```
In the last model, we assumed ignorance about the probability of seeing frogs at the pond, by using an uninformed uniform prior.
What if we had some prior knowledge about this pond? For example, we know that this pond is in an area where frogs are very common, and we expect to see them most of the time.
We can use a Beta distribution to represent this prior knowledge.
Why a Beta distribution? Because it is defined on the interval [0, 1], which is suitable for probabilities, and it can take on a variety of shapes depending on its parameters.

But what parameters should we use for the Beta distribution? Let's say we expect to see frogs 80% of the time, and we are quite confident about this.
We can use the parameters alpha = 5 and beta = 8, which gives us a prior mean of 0.4 and a relatively narrow distribution.

We can also plot this prior distribution to see how it looks.
```{r}
plot(function(x) dbeta(x, 5, 8), from=0, to=1, col="orange", ylab="Density", xlab="Probability of seeing frogs")
print(paste("Mean of Beta(5,8):", 5/(5+8)))
print(paste("Mode of Beta(5,8):", (5-1)/((5-1)+(8-1))))
```
And what we can see is that this prior distribution is centered around 0.4, and it has a peak at this value.
You can play around with the shape values to move the peak to center the distribution around different values.
Now, let's fit the model again using this informed Beta prior.
```{r}
# MCMC method using ulam with informed Beta prior
frog_model_beta <- ulam(
  alist(
    frogs ~ dbinom(trips, p), # data likelihood
    p ~ dbeta(a = 5, b = 8) # p prior
  ), data = data_list
)

```

```{r}
# model summary
precis(frog_model)
```

```{r}
# posterior samples
samples <- extract.samples(frog_model, n=10000)

# plots
curve(dnorm(x, mean(samples$p), sd(samples$p)), xlab="Probability of seeing frogs", ylab="Density", col="blue")
curve((dunif(x, 0, 1)), add = TRUE, col="orange")
abline(v = 0.35, col="red")
legend("topright", legend=c("Posterior", "Prior", "True Probability"),
       col=c("blue", "orange", "red"), lty=1)
```

```{r}
# posterior summaries
# above 35%
sum(samples$p > 0.35) / length(samples$p)

# compatibility intervals
# percentage interval
PI(samples$p, prob=0.5)
# highest posterior density interval
HPDI(samples$p, prob=0.96)
```
```{r}
# plot HDPI
dens(samples$p, xlab="Probability of seeing frogs", ylab="Density")
abline(v = 0.35, col="red")
abline(v = HPDI(samples$p, prob=0.96), col="blue", lty=2)
legend("topright", legend=c("True Probability", "96% HPDI"),
       col=c("red", "blue"), lty=c(1,2))
```
